---
title: "OncoScan: Mutation-Outcome Association Explorer in Breast Cancer (METABRIC)"
author: "Kennedy Kasonde"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(ggrepel)
```

# Introduction

Breast cancer is a molecularly heterogeneous disease, and mutations in key
driver genes are known to associate with distinct clinical phenotypes and
prognosis. This report explores somatic mutation data from the METABRIC
cohort (Curtis et al., 2012; Pereira et al., 2016) to characterize mutation
frequency across a curated oncogene panel and test whether mutation status
in these genes is associated with the Nottingham Prognostic Index (NPI), a
composite clinical score combining tumor size, grade, and nodal status.

This analysis is exploratory and statistical in nature — it does not
involve machine learning or predictive modeling. All data wrangling uses
the `tidyverse` (`dplyr`, `stringr`, `forcats`), and all statistical
testing uses base R with multiple-testing correction via
`p.adjust(method = "BH")`.

**Data source:** METABRIC cohort, accessed via cBioPortal
(https://www.cbioportal.org/study/summary?id=brca_metabric).

---

# Data Loading and Cleaning

```{r load-data}
Data <- read_csv("data/METABRIC_RNA_Mutation.csv")
glimpse(Data)
```

## Clinical data preparation

We select the clinical variables relevant to this analysis and derive a
triple-negative breast cancer (TNBC) status flag from receptor status
columns. `death_from_cancer` contains blank strings rather than proper
`NA`s in the raw export, so these are converted explicitly before further
use.

```{r clean-clinical}
Data$death_from_cancer <- na_if(Data$death_from_cancer, "")

clinical_data <- Data %>%
  select(patient_id, age_at_diagnosis, tumor_size, tumor_stage,
         er_status, pr_status, her2_status,
         overall_survival_months, overall_survival, death_from_cancer,
         nottingham_prognostic_index, mutation_count,
         chemotherapy, hormone_therapy, radio_therapy) %>%
  mutate(
    triple_negative_status = case_when(
      er_status == "Negative" & pr_status == "Negative" & her2_status == "Negative" ~ "TNBC",
      TRUE ~ "Non-TNBC"
    )
  )

glimpse(clinical_data)
```

## Oncogene panel definition

A 15-gene panel of well-characterized breast cancer oncogenes and tumor
suppressors was selected and verified against the dataset's column names
before use.

```{r oncogene-panel}
oncogene_panel <- c("tp53", "brca1", "brca2", "pik3ca", "kras", "braf",
                     "egfr", "chek2", "akt2", "cdkn2a", "gata3", "cdh1",
                     "map3k1", "erbb2", "pten")

mut_cols <- paste0(oncogene_panel, "_mut")
stopifnot(length(setdiff(mut_cols, colnames(Data))) == 0)
```

## Mutation data reshaping

Mutation columns are stored in wide format, with `"0"` indicating no
mutation and any other value containing the protein-change annotation
(e.g., `"R175H"`). We reshape to long format and classify each mutated
entry by variant type using `str_detect()` on the annotation text.

**Note on scope:** a small number of cells (~240 of 2,344 mutated
instances, ~10%, concentrated in PIK3CA and MAP3K1) contain multiple
co-occurring mutations in the same gene. These are classified by the
first-matching pattern in the `case_when()` below rather than split into
separate events — a documented simplification (see Limitations).

```{r reshape-mutations}
mutation_long <- Data %>%
  select(patient_id, all_of(mut_cols)) %>%
  pivot_longer(
    cols = all_of(mut_cols),
    names_to = "gene",
    values_to = "mutation_detail"
  ) %>%
  mutate(
    gene = str_remove(gene, "_mut"),
    mutated = mutation_detail != "0",
    mutation_type = case_when(
      mutation_detail == "0" ~ "None",
      str_detect(mutation_detail, "fs") ~ "Frameshift",
      str_detect(mutation_detail, "_splice") ~ "Splice site",
      str_detect(mutation_detail, "\\*") ~ "Nonsense",
      str_detect(mutation_detail, "del") ~ "In-frame deletion",
      TRUE ~ "Missense"
    )
  )
```

---

# Results

## Mutation frequency across the oncogene panel

```{r gene-frequency}
gene_freq <- mutation_long %>%
  filter(mutated) %>%
  count(gene, name = "n_mutated", sort = TRUE) %>%
  mutate(
    pct_patients = n_mutated / nrow(clinical_data) * 100,
    gene = str_to_upper(gene)
  )
```

```{r frequency-plot, fig.width=9, fig.height=6, fig.cap="Figure 1: Mutation frequency across the METABRIC oncogene panel."}
ggplot(gene_freq, aes(x = fct_reorder(gene, n_mutated), y = pct_patients)) +
  geom_col(fill = "steelblue") +
  geom_text(aes(label = round(pct_patients, 1)), hjust = -0.1) +
  coord_flip() +
  labs(
    title = "Mutation Frequency Across METABRIC Oncogene Panel",
    x = "Gene Name", y = "% of Mutated Patients",
    caption = "Data Source: METABRIC Cohort"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "italic")
  )
```

PIK3CA (`r round(gene_freq$pct_patients[gene_freq$gene == "PIK3CA"], 1)`%)
and TP53 (`r round(gene_freq$pct_patients[gene_freq$gene == "TP53"], 1)`%)
are the most frequently mutated genes in this cohort, consistent with
published METABRIC mutation frequencies (Pereira et al., 2016). Genes such
as KRAS, BRAF, and AKT2 show very low mutation frequency here, consistent
with their more prominent driver role in other cancer types rather than
breast cancer specifically.

## Clinical stratification

```{r clinical-plot-data}
clinical_plot_data <- clinical_data %>%
  filter(!is.na(tumor_stage)) %>%
  mutate(tumor_stage = factor(tumor_stage))
```

```{r tumor-size-plot, fig.width=7, fig.height=5, fig.cap="Figure 2: Tumor size by TNBC status."}
ggplot(clinical_plot_data, aes(x = triple_negative_status, y = tumor_size,
                                fill = triple_negative_status)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 1.5) +
  labs(
    title = "Distribution of Tumor Size by TNBC Status",
    x = "Triple-Negative Breast Cancer (TNBC) Status", y = "Tumor Size (mm)",
    caption = "Data Source: METABRIC Cohort"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        legend.position = "none")
```

```{r npi-stage-plot, fig.width=7, fig.height=5, fig.cap="Figure 3: NPI across tumor stages."}
ggplot(clinical_plot_data, aes(x = tumor_stage, y = nottingham_prognostic_index,
                                fill = tumor_stage)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 16, outlier.size = 1.5) +
  labs(
    title = "Nottingham Prognostic Index (NPI) Across Tumor Stages",
    x = "Tumor Stage", y = "Nottingham Prognostic Index (NPI)",
    caption = "Data Source: METABRIC Cohort"
  ) +
  theme_classic(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14, hjust = 0.5),
        legend.position = "none")
```

NPI increases consistently with tumor stage from Stage 0 through Stage 3.
Stage 4 (n = `r sum(clinical_plot_data$tumor_stage == "4")`) shows a lower
median NPI than Stage 3; given the very small sample size at this stage,
and that NPI does not directly incorporate metastatic spread (unlike
overall tumor staging), this pattern should be interpreted cautiously
rather than as a contradiction.

## Statistical association: mutation status and prognostic score

For each gene in the panel, patients were split into mutated vs.
non-mutated groups and compared on NPI using Welch's t-test. Raw p-values
were adjusted across all 15 tests using the Benjamini-Hochberg procedure
to control the false discovery rate.

```{r join-mutation-clinical}
mutation_clinical <- mutation_long %>%
  filter(gene %in% oncogene_panel) %>%
  select(patient_id, gene, mutated) %>%
  left_join(
    clinical_data %>% select(patient_id, nottingham_prognostic_index),
    by = "patient_id"
  )
```

```{r gene-association-tests}
gene_results <- mutation_clinical %>%
  group_by(gene) %>%
  group_modify(~ {
    npi_mut    <- .x$nottingham_prognostic_index[.x$mutated]
    npi_nonmut <- .x$nottingham_prognostic_index[!.x$mutated]

    test_result <- tryCatch(
      t.test(npi_mut, npi_nonmut),
      error = function(e) NULL
    )

    if (is.null(test_result)) {
      tibble(n_mutated = sum(.x$mutated), mean_diff = NA_real_, p_value = NA_real_)
    } else {
      tibble(
        n_mutated = sum(.x$mutated),
        mean_diff = mean(npi_mut) - mean(npi_nonmut),
        p_value   = test_result$p.value
      )
    }
  }) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_value, method = "BH"))

gene_results %>% arrange(p_adj)
```

```{r volcano-data}
volcano_data <- gene_results %>%
  mutate(
    neg_log10_p = -log10(p_adj),
    significant = p_adj < 0.05,
    gene_label = str_to_upper(gene)
  )
```

```{r volcano-plot, fig.width=10, fig.height=6, fig.cap="Figure 4: Association between oncogene mutation status and Nottingham Prognostic Index."}
ggplot(volcano_data, aes(x = mean_diff, y = neg_log10_p, color = significant)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_text_repel(
    data = filter(volcano_data, significant),
    aes(label = gene_label),
    fontface = "bold", show.legend = FALSE,
    box.padding = 0.6, point.padding = 0.4,
    min.segment.length = 0, segment.color = "grey40", force = 3
  ) +
  scale_color_manual(values = c("TRUE" = "firebrick", "FALSE" = "grey60")) +
  labs(
    title = "Association Between Oncogene Mutation Status and Prognostic Score",
    subtitle = "Nottingham Prognostic Index by mutation status across 15 breast cancer oncogenes (METABRIC)",
    x = "Mean Difference in NPI (Mutated − Non-mutated)",
    y = expression(-log[10](adjusted~italic(p)~value)),
    color = "Significant\n(BH-adjusted p < 0.05)",
    caption = "Data Source: METABRIC Cohort | Benjamini-Hochberg correction applied across 15 tests"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold", size = 14), legend.position = "right")
```

Of the 15 genes tested, `r sum(volcano_data$significant)` remained
significant after BH correction: TP53, GATA3, PIK3CA, and CDH1. TP53
mutation is associated with a *higher* NPI (mean difference =
`r round(volcano_data$mean_diff[volcano_data$gene == "tp53"], 2)`),
consistent with its role as a tumor suppressor whose loss is linked to
more aggressive disease. GATA3, PIK3CA, and CDH1 mutations are each
associated with *lower* NPI, consistent with their association with
lower-grade, often luminal or lobular breast cancer subtypes (Curtis et
al., 2012).

It is worth noting that statistical significance here does not
necessarily imply large clinical significance — for example, the TP53
effect size (~0.7 points on a scale from roughly 2 to 6.7) is modest
relative to its extremely small p-value, which is driven substantially by
the large sample size (n = `r nrow(clinical_data)`).

---

# Limitations

- `tumor_stage` is missing for `r sum(is.na(Data$tumor_stage))` patients
  (~26% of the cohort); analyses using this variable reflect only the
  staged subset.
- `tumor_size` is missing for `r sum(is.na(Data$tumor_size))` patients.
- Patients with multiple mutation events in the same gene (~10% of
  mutated instances, concentrated in PIK3CA and MAP3K1) were classified
  by a single dominant mutation type rather than split into separate
  events.
- Association testing uses `overall_survival_months`'s companion metric,
  NPI, rather than time-to-event survival analysis, since the latter
  requires handling censoring (patients still alive at last follow-up),
  which is outside the current scope.
- This analysis is exploratory and hypothesis-generating; it does not
  establish causality between mutation status and prognostic score.

# Future Improvements

- Proper survival analysis (Kaplan-Meier, Cox proportional hazards) using
  `overall_survival_months` with censoring handled correctly.
- Splitting multi-mutation cells into individual variant-level rows.
- Extending the association analysis to gene expression z-scores
  alongside mutation status.
- Incorporating molecular subtype (PAM50) as a covariate or stratification
  variable.

# References

- Curtis, C. et al. (2012). The genomic and transcriptomic architecture of
  2,000 breast tumours reveals novel subgroups. *Nature*, 486, 346–352.
- Pereira, B. et al. (2016). The somatic mutation profiles of 2,433 breast
  cancers refine their genomic and transcriptomic landscapes. *Nature
  Communications*, 7, 11479.
- Data accessed via cBioPortal for Cancer Genomics
  (https://www.cbioportal.org/study/summary?id=brca_metabric).
